---
title: "Homework on estimating dynamic discrete choice models"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float:
      collapsed: false
      smooth_scroll: false
---

```{r,echo=TRUE,message=FALSE,warning=FALSE}
library(data.table)
library(mvtnorm)
library(ggplot2)
library(futile.logger)
library(pander)
```

In this homework we want to estimate a simple dynamic discrete choice model by Keane and Wolpin that we saw in class. They propose a simple model of female labor supply as follows. Utility for married woman $i$ in period $t$ from working (option 1) vs not working (option 0) with $n_i$ small children is

$$
\begin{align}
U_{it}^1 &= y_{it} + w_{it} - \pi n_{it} \\
U_{it}^0 &= y_{it} + x_{it}\beta + \epsilon_{it} 
\end{align}
$$
where $y_{it}$ is the husband's income, and *importantly*, $n\subset x$, i.e. you get utility from children while not working. Let's write the difference in those utilities as $U_{it}^1 - U_{it}^0$,

$$
v_{it}(x_{it},w_{it},n_{it},\epsilon_{it}) = w_{it} - \pi n_{it} -x_{it}\beta - \epsilon_{it} 
$$
and define the work indicator $d_{it} = \mathbf{1}[U_{it}^1 > U_{it}^0]$. Next, we define the state space as observed by the econometrician as $\Omega_{it} = (x_{it},w_{it},n_{it})$, i.e. we don't observe $\epsilon$, but the decision maker does.

Woman $i$ will work in $t$ if $U_{it}^1 > U_{it}^0$, i.e. if $v_{it}(x_{it},w_{it},n_{it},\epsilon_{it}) > 0$, and at $v_{it}(x_{it},w_{it},n_{it},\epsilon_{it}) = 0$ she is indifferent. Call the $\epsilon$ that solves this *the critical epsilon* $\epsilon^*(\Omega_{it})$. We have that, given state $\Omega_{it}$,

$$
i\text{ chooses to }\begin{cases}\text{work in t}&\text{if }\epsilon_{it} < \epsilon^*(\Omega_{it}) \Rightarrow U_{it}^1 > U_{it}^0 \\
\text{not work in t}& \text{if }\epsilon_{it} > \epsilon^*(\Omega_{it}) \Rightarrow U_{it}^1 < U_{it}^0\end{cases}
$$

Assuming that $\epsilon$ is independent of $\Omega$, the probability of working is computed by looking at each $i$'s $\epsilon_{it}$ and counting how many are below $\epsilon^*(\Omega_{it})$:

$$
\Pr[d_{it}=1|\Omega{it}] = \int_{-\infty}^{\epsilon_{it}^*} dF_{\epsilon}(\epsilon|\Omega{it}) = \int_{-\infty}^{\epsilon_{it}^*} dF_{\epsilon}(\epsilon)
$$

Clearly $\Pr[d_{it}=0|\Omega{it}] = 1- \Pr[d_{it}=1|\Omega{it}]$, and so the likelihood for a random sample of $N$ females observed for $T$ periods is

$$
L(\beta,\pi,F_{\epsilon};x_{it},w_{it}) = \Pi_{i=1}^N \Pi_{t=1}^T \Pr[d_{it}=1|\Omega{it}]^{d_{it}} \Pr[d_{it}=0|\Omega{it}]^{1-d_{it}}
$$

# Simulating a data set

We start out with a simulated data set that we will use in estimation.

```{r,echo=TRUE,message=FALSE,warning=FALSE}
# true parameters
p = list()
p$beta1 = 1
p$beta2 = 0.5
p$beta3 = 0.5
p$pi = 1.5
p$N = 2000
p$T = 60  # final year of observation

get.data <- function(p){
  d = data.table(expand.grid(id = 1:p$N,it=20:p$T))
  d[,n := sample(c(0,1,2),size=1), by=id]
  d[,x1 := runif(n=nrow(d))]
  d[,x2 := runif(n=nrow(d))]
  d[,w := rnorm(nrow(d),mean=5)]  # suppose we observe wages of non-workers
  d[,eps := rnorm(mean=0,sd=1,n=nrow(d))]
  d[,v:= w - p$pi*n - p$beta1*x1- p$beta2*x2 - p$beta3*n -eps]
  d[,work := v > 0]
  flog.info("participation rate: %f",d[,mean(work)])
  return(d)
}
d <- get.data(p)
```


## R-golfing

<span class="btn btn-success btn-xs ">Question 1.</span> Write a one-liner that computes the means of x,w,eps,v and work from this data.table. count how many characters you had to type with `nchar("YOUR_CODE_IN_QUOTES_LIKE_THIS")`. I have 32.


## Estimating a static model

<span class="btn btn-success btn-xs ">Question 2.</span> Identification.

1. Run a regression of `v` on `w,x,n` and see if you can recover the true parameters.
1. Make suitable assumptions to be able to recover all parameters.


<span class="btn btn-success btn-xs ">Question 3.</span> Functional Forms.

1. What does the likelihood function above remind you of?
2. Sometimes people use a linear probability model to approximate a probit model.  Which parameters can we recover using an LPM here? How well does it predict the average probability of work? Give the quartiles of the predicted probabilities. 
3. Then do the same exercise with a probit model.


Let's make the wage a bit more realistic. Suppose we knew completed years of education, and we had a variable $z_1$ capturing *potential experience* (age - educ - 6), which determines the wage, $z_2 \sim  U(0,1)$ as a second explanatory variable  and $\eta$ a random error. Additionally, let's be more reasonable and assume that for all non-workers, we don't observe a wage, for which we code `0`. 

$$
w_{it} = z_{it}'\gamma  + \eta_{it}
$$

We posit that $\eta$ and $\epsilon$ are joint normal with mean zero and var-cov matrix

$$
\sum = \left[ \begin{array}{cc}
\sigma_\eta^2 & \sigma_{\eta \epsilon} \\
\sigma_{\eta \epsilon}& \sigma_\epsilon^2 
\end{array} \right] = 
\left[\begin{array}{cc}
0.5  & 0.5 \\
0.5 & 0.8 
\end{array}
\right]
$$

<span class="btn btn-success btn-xs ">Question 4.</span> Change the function `get.data` to reflect those changes. Add a new parameters $\gamma_1=1,\gamma_2=2$ to `p`. Then run a wage regression and see if you can recover the $\gamma$s. Report the value of $E[\eta|\text{work}]$ and of $E[\eta]$.

> In my solution, I had big problems getting identification with the so-specified model. I therefore changed that part of the model so that $z_1 \sim U[-5,5], z_2 \sim U[-5,5]$. This increases the identifiability of the likelihood function a lot, as you will see further below

<span class="btn btn-success btn-xs ">Question 5.</span> Estimate the static model with maximum likelihood.  

1. write a likelihood function that takes our simulated dataset `d`, the contents of `p` and returns the value of a the likelihood function above (i.e. it returns a number)

```{r}
# the way you set up this function depends a lot on the optimizer you are going to use later on. We will be using maxLik.
static.like <- function(beta1,beta2,beta3,gamma1,gamma2,var_eps,var_eta,cov,data){
    
  # enforce non-negativity constraint on parameters.
  # some optimizers allow to have box constraints on choice variables,
  # which would be preferrable. Here, in the absence of such functionality,
  # this is an alternative solution.
  
  # compute \xi
  
  # compute probability of no work
 

  # joint probability of work and certain wage
  # pr(xi > -v and eta = w - z \gamma) =
  # pr(eta-eps > -v and eta = w - z \gamma) =
  # pr( eps < w - xb )

  
  # compute value of log likelihood for each individual

  
  # return sum of log of likelihood


}
# test run
x=static.like(p$beta1,p$beta2,p$beta3,p$gamma1,p$gamma2,p$var_eps,p$var_eta,p$cov,d)

```

2. check the likelihood function: produce a plot that has a panel for each parameter, and each panel shows the value of the likelihood function as the parameter varies by +/- 20% from it's starting value in `p` (while one parameter varies, the others are fixed at their value in `p`).

```{r,echo=FALSE}
# wrapper 1: a closure to include the dataset (can't pass as parameter to optimizer!)
dwrap.like <- function(beta1,beta2,beta3,gamma1,gamma2,var_eps,var_eta,cov){
    #static.like(beta1,beta2,beta3,gamma1,gamma2,var_eps,var_eta,cov,d)
}

# wrapper 2: maps a numeric vector to function arguments.
mwrap.like <- function(x){
    # static.like(beta1   =x[1],
    #             beta2   =x[2],
    #             beta3   =x[3],
    #             gamma1  =x[4],
    #             gamma2  =x[5],
    #             var_eps =x[6],
    #             var_eta =x[7],
    #             cov     =x[8],
    #             data    = d)
}

# now check that the likelihood is correct.
# make a grid of values for each param +- 20 % of starting value
n = 50  # number of points in each panel
l = 0.2

```

3. Set up an optimizer that searches for a maximum over that function. Fix $\gamma_2$.

```{r}
require(maxLik)
# could constrain var_eps and var_eta to be positive
# A <- matrix(c(rep(0,5),1,0,0,
#               rep(0,5),0,1,0), 2, 8,byrow=T)
# B <- c(0,0)
# mle = maxLik(mwrap.like,grad=NULL,hess=NULL,start=unlist(l1) + runif(length(l1))*0.1,fixed=c("gamma2"),control=list(printLevel=1),constraints=list(ineqA=A,ineqB=B))
# prefer setting mwrap.like=NA if negative.
# mle = maxLik(mwrap.like,grad=NULL,hess=NULL,start=unlist(l1) + runif(length(l1))*0.1,fixed=c("gamma2"),control=list(printLevel=1),method="BFGS")
# summary(mle)
```

<span class="btn btn-success btn-xs ">Question 6.</span> Do a counterfactual experiment with this model. Illustrate how labor supply varies as we introduce a child care subsidy of $\tau=1$ dollars if the woman works while $n_{it}>0$. 

* You can use the *true* parameter values stored in `p`. 
* Ideally you would show a plot with two overlaid density estimates. 
    1. A kernel estimate of the distribution of $\Pr(d_{it}=1|\Omega_{it},\tau=0)$
    2. A kernel estimate of the distribution of $\Pr(d_{it}=1|\Omega_{it},\tau=1)$
* Also report a table with mean and median of that distribution by scenario.
    

<span class="btn btn-success btn-xs ">Question 7.</span> Add experience ($h$) to our dataset and to the wage equation. The law of motion of $h$ is

$$
h_{it} = h_{it-1} + d_{it-1}
$$
i.e. $h$ increases by 1 if the individual worked last period.


<span class="btn btn-success btn-xs ">Question 8.</span> Check the created dataset with at least those two conditions:

1. everybody starts with 0 experience
1. `work = v > 0` everywhere

```{r}
# library(testthat)
# test_that("everybody starts with 0 experience", {
#     expect_true(all(d[it==min(it),h==0]))
# })
# test_that("work = v > 0", {
#     expect_true(all(d[,work == (v > 0)]))
# })
```

## Estimating a Dynamic Model

Now that $h$ has been added, our model has finally become dynamic: *Tomorrow's value depends on what you do today*. Here, tomorrow's wage depends on whether you worked in all previous periods, hence trading off the decision of whether or not to work has a component that describes the future consequences of each choice.

<span class="btn btn-success btn-xs ">Question 9.</span> solve the dynamic model.  


<span class="btn btn-success btn-xs ">Question 10.</span> estimate dynamic model







