---
title: "Value Function Iteration Example"
output:
  html_document:
    df_print: paged
---

## Deterministic Growth Model Dynamic Program

(This is my version of the example at [Sargent and Stachurski's quant-econ website](quant-econ.net/jl/optimal-growth.jl). Please observe the license file at the root of that repository.)

In this notebook we'll implement the deterministic growth model as a dynamic programming problem. We will assume log utility to get a closed form solution. 
Remember that the problem is defined as
\begin{align}
   V(k) &= \max_{0<k'<f(k)} \ln(f(k) - k') + \beta V(k')\\
     f(k)   & = k^\alpha\\
     k_0   & \text{ given} 
\end{align}

## Representing a function on $\mathbb{R}$ in a computer
     
* The first thing to realise is that we cannot represent $V(k),k\in \mathbb{R}$ on a computer.  However, we can get an arbitrarily good approximation to  $\mathbb{R}$. 
* We will approximate $k$ at a finite number of points $K={k_1,\dots,k_n}$, called a *grid*. 
* In other words, we will compute $V(k)$ above only at the list of points in $K$. 
* There is a slight complication that arises from the $\max$ operator: 
    * Ideally, we would like to choose $c$ out of the **continuous** interval $[0,f(k)]$, and not be restricted to the grid $K$. 
    * In order to achieve this, we must find a way to represent $V(k)$ for values off the grid. 
    * In other words, we will know a list of values $V(k_1),\dots,V(k_n)$, but will most of the time need a value $V(x),x\in (k_i,k_{i+1})$ when we perform the operation $\max_{0<k'<f(k)}$. 
    * We will *linearly interpolate* such a value $x$, which is similar to *connecting the dots*.

## Fire up **R**

```{r}
alpha     = 0.65
beta      = 0.95
grid_max  = 2  # upper bound of capital grid
n         = 150
kgrid     = seq(from=1e-6,to=grid_max,len=n)  # equispaced grid
f <- function(x,alpha){x^alpha}  # defines the production function f(k)
```

Next, because of our log assumption, we know that there is a closed form solution here. It is characterized by 2 constants $c_1,c_2$. We know the true solution to the value function, denoted $V^*$:

```{r}
ab        = alpha * beta
c1        = (log(1 - ab) + (log(ab) * ab / (1 - ab))) / (1 - beta)
c2        = alpha / (1 - ab)
v_star <- function(k,c1,c2){c1 + c2 * log(k)}	# this defines a function v_star
```


We will now apply the bellman operator to the functional in the above definition. The operator takes a current guess $V^i$ and returns the next iterate $V^{i+1}$. We define the operator as
\begin{align} T(V)(k) =& \max_{0<k'<f(k)} \ln(f(k) - k') + \beta V(k') \\
                V^{i+1}(k) =& \max_{0<k'<f(k)} \ln(f(k) - k') + \beta V^{i}(k') 
\end{align}

```{r}
bellman_operator <- function(grid, w){
    # 1) we need an object that interpolates the current guess in w
    Interp = approxfun(grid, w)
    
    # 2) create a vector to hold the result
    Tw = rep(0,length(w))

    # 3) for all grid points k, do the maximization
    for (i in 1:length(grid)){
        k = grid[i]
        # 4) at each grid point, define an objective function
        objective <- function(c){ - log(c) - beta * Interp(f(k,alpha) - c)}
        # 5) and find the max of that function. 
        # find max of ojbective between [0,k^alpha]
        res = optimize(objective, lower=1e-6, upper=f(k,alpha)) 
        # 6) save that in the result vector
        Tw[i] = - objective(res$minimum)
   }
    #7) return the next guess
    return(Tw)
}
```

Now we can start the value function iteration:

```{r}
# value function iteration (VFI) 
# input:  * Integer maxIter: max number of iterations
# output: * matrix Vfuns: each column is an iterate on V
VFI <- function (grid,V0,maxIter){
    w = matrix(0,length(grid),maxIter)
    w[,1] = V0 # initial condition
    # start iteration
    for (i in 2:maxIter){
        w[ ,i] = bellman_operator(grid, w[ ,i-1])
    }
    return(w)
}
```

### Let's run this now!

```{r}
# let's do it!
v0 = 5 * log(kgrid) - 25  #initial condition
V = VFI(kgrid,v0,35)

# plot it
truth = data.frame(grid = kgrid, v=v_star(kgrid,c1,c2),variable = 0)
d = as.data.frame(V)
d$grid = kgrid
library(reshape2)
m = melt(d,id.vars = "grid")

library(ggplot2)
p <- ggplot(m,aes(x=grid,y=value,color=as.integer(variable),group=variable)) + 
    geom_line() +
    scale_y_continuous(limits = c(-50,-20)) +
    scale_colour_gradient(name = "iteration", low = "grey", high= "black")
p <- p + geom_line(data=truth,aes(x=grid,y=v,group=variable),color="red") 
p
```

Ok that seems to work. What about a random starting value?

```{r}
v0 = runif(length(kgrid)) - 25  #initial condition
V = VFI(kgrid,v0,70)
d = as.data.frame(V)
d$grid = kgrid
m = melt(d,id.vars = "grid")
p2 <- ggplot(m,aes(x=grid,y=value,color=as.integer(variable),group=variable)) + 
    geom_line() +
    scale_y_continuous(limits = c(-50,-20)) +
    scale_colour_gradient(name = "iteration", low = "grey", high= "black")
p2 + geom_line(data=truth,aes(x=grid,y=v,group=variable),color="red") 
p2
```
